---
title: "ENVX2001 - Model selection"
# revised-by: aca
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Learning Outcomes

In this lab, you will work towards achieving learning outcomes

## Lab Objectives

In this lab, we will:

- []
- []

:::{.callout-tip}
Please work on this exercise by creating your own R Markdown file.
:::


## Exercise 1: Model quality - multiple vs adjusted *r*^2^ 

Data:  *California_streamflow* spreadsheet  

Import the "California_streamflow" sheet into R.

```{r}
# Load library if needed
library(readxl)
# Load data
stream_data <- read_xlsx("data/california_streamflow.xlsx", "streamflow")
```

in this exercise we will use the same data as last week. To jog your memory, the dataset contains 43 years of annual precipitation measurements (in mm) taken at (originally) 6 sites in the Owens Valley in California. Through model selection via partial F-test we have the final model as below:

```{r}
fit <- lm(runoff_volume ~ rock_creek + pine_creek, data = stream_data)
```

We will now add a totally useless variable to the dataset. This variable is a random number generated from a normal distribution with mean 3 and standard deviation 2. We use the `set.seed()` function to make sure that everybody gets the same random values.

```{r}
set.seed(100) # to make sure everybody gets the same results

# this generates the random number into the dataset
stream_data$random_no <- rnorm(n = nrow(stream_data), mean = 3, sd = 2) 
```
  
We will see the impact of including a totally useless variable, such as this random variable, has on measures of model quality, *r*^2^ and adjusted *r*^2^ values.

**Task: create two regression models:**

1.  `runoff_volume ~ rock_creek + pine_creek` 
2.  `runoff_volume ~ rock_creek + pine_creek + random_no`  

:::{.question}
### Question 2
Compare each in terms of their multiple *r*^2^ and adjusted *r*^2^ values. Which performance measure (multiple *r*^2^ or adj *r*^2^) would you use to identify which predictors to use in your model?  
:::

::: {.content-visible when-profile="solution"}
### Answer {style="color:green;"}

There is only a small difference in the multiple *r*^2^ and adj *r*^2^ between the models, but one goes up and the other goes down. 
Since the random number is a totally useless value, this demonstrates that the adjusted *r*^2^ is the better performance measure to use. 


```{r randno_solution, eval = T, echo = T}
mod_rand1 <- lm(runoff_volume ~ rock_creek + pine_creek,data = stream_data)
summary(mod_rand1)

mod_rand2 <- lm(runoff_volume ~ rock_creek + pine_creek + random_no ,data = stream_data)
summary(mod_rand2)
```

:::
## Exercise 2: Fish productivity

Fish communities were surveyed in lakes across a eutrophication gradient to investigate the relatioship between productivity and fish diversity.

The datasheet has thefollowing variables:

- `lake_id` Unique identifier for each lake 
- `chla` Log-transformed Chlorophyll a
- `richness` Log-transformed species richness
- `evenness` Log-transformed Pielou’s evenness 	
- `abundance` Log-transformed abundance per unit effort (NPUE) 	
- `biomass` Log-transformed biomass per unit effort (BPUE) 
- `productivity` Log-transformed productivity proxy

```{r}
#Load library if necessary
library(tidyverse)

#Read in data
fish_data <- read_csv("data/fish_communities.csv")

#Have a look at the structure of the data
str(fish_data)
```

Explore the data on your own before making any models. 
(*Hint:* remember that we are only interested in the numeric variables for our linear models)

:::{.question}
### Question 1

Are there any obvious relationships between the response variable (productivity) and the other variables?
:::

::: {.content-visible when-profile="solution"}
### Answer {style="color:green;"}

```{r}
#correlation matrix
cor(fish_data[,-1])

#Plot the variables

pairs(fish_data[,-1])
```
Productivity has a strong relationship with biomass and abundance
:::

:::{.question}
### Question 2

Are there any other patterns or potential issues you can see from the exploratory plots?
:::

::: {.content-visible when-profile="solution"}
### Answer {style="color:green;"}

Biomass is also highly correlated with abundance, so collinearity might be an issue.
:::


## Take home exercise:

## Review

### Attribution
